---
title: "Using R Markdown Files"
author: "Michael Gavin"
date: "January 19"
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
---
For our work in R, I'll communicate with you using R Markdown files. These will be kind of weird to work with at first, but, trust me, they make everything a lot easier. The nice thing about Markdown files is that you can embed code right into them. Which means I can give you examples, and you can just click on little green arrow to run the code. When you execute code within the notebook, the results appear beneath the code. 

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*. 

```{r}
print("Question = 2B | !2B")
```

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

Throughout the semester, we'll be using a number of R packages, which come with built-in functions. Install and load them.
```{r}
install.packages("Matrix")
install.packages("igraph")
install.packages("xml2")
install.packages("devtools")
```

Of these, for today we'll be using `devtools` and `Matrix`. This command activates their libraries of functions.
```{r}
library(Matrix)
library(devtools)
```

All data for this course is contained in a package called `math4shakespeare`, but it's a little different from the others, because it's not an official R package. It's just stored on Github. You need to use `devtools`, which has a special function called `install_github()` to install it.
```{r}
install_github("michaelgavin/math4shakespeare")
```

Now we can look at the datasets we'll be using throughout the semester. **P** is a words x plays matrix. You can get its dimensions like this:
```{r}
library(math4shakespeare)
data("P")
dim(P)
```

The `dim()` function tells you how many rows and columns are in the matrix. The matrix **P** has 26,952 rows (for each unique word form) and 37 columns (for each play).

To calculate the total frequencies for each word, take the sum of each row:
```{r}
totals = rowSums(P)
```

Now look at the top ten words
```{r}
sort(totals, decreasing = T)[1:10]
```

S is a words x speeches matrix.
```{r}
data("S")
dim(S)
```

You'll notice it's much bigger because it has a lot more columns (one for every speech given by every character in every play), but it has the same number of rows.
```{r}
totals = rowSums(S)
```

Now check this out:
```{r}
sort(totals, decreasing = T)[1:10]
```

The values should be exactly the same. Why? Because it's all the same data, organized into different columns. **P** just has the totals for each play. **S** has each speech. And **C** has the word counts by character.


```{r}
data("C")
dim(C)
```

There are 1448 characters across the 37 plays, plus an extra column for undefined speeches. (That happens rarely when several characters are shouting simultaneously, and the editors did not assign a speaker to the speech.)

Now check this out:
```{r}
sum(P)
sum(S)
sum(C)
```

If the data is correct, these totals should all match exactly -- again, because it's the same textual data, just divided differently.

Above, we used `rowSums()` to get the total frequency for each word. Now let's get the total for each character to see who talks the most.
```{r}
totals = colSums(C)
sort(totals, decreasing = T)[1:5]
```

This makes sense. The history plays are really long, and some have sequels. So Henry V and Falstraff are both in two plays. 

Also, Richard III and Hamlet talk a lot.

### Analyzing word counts

Check your data, first by creating a variable x
```{r}
x = S["death",]
```

Then see how many x's are > 0.
```{r}
length(x[x>0])
```

If the answer is 716, you're doing well. What that means is that there are 716 speeches in all of Shakespeare's plays that use the word "death" at least once.

Notice that the total frequency is higher:
```{r}
sum(x)
```

**Comprehension check:** Why is `length(x[x>0])` less than `sum(x)`?

If you can't answer that question, stop and re-read the above and think about what these two functions are counting.

You can look at the play metadata
```{r}
data("play_metadata")
View(play_metadata)
```

If you play around with the `View()` function, you'll see that you can sort by different columns. For small data, it's really useful. But don't try to `View()` a large dataset with thousands of columns. You'll crash RStudio.

And you can select from the data, too. (This will be important later.)
```{r}
View(play_metadata[play_metadata$GENRE == "HISTORY",])
```

To get the number of words spoken in history plays, use the metadata to create an index (which I'm in the habit of calling `hits`) that subsets from **P** as a whole.
```{r}
hits = which(play_metadata$GENRE == "HISTORY")
ids = play_metadata$ID[hits]
totals = rowSums(P[,ids])
totals["death"]
```

Of the 837 times Shakespeare used the word "death," 344 were in history plays.

**Comprehension check:** What is `ids` in the above code? What does `P[,ids]` refer to?

### Working with character metadata

You can look at the metadata for each character:
```{r}
data("character_metadata")
View(character_metadata)
```

To get the list of male characters, create an index, just like above. (In my code, any time I do a search, I just save the results of that search in a variable that I call `hits`. I don't store the results of the search permanently; I re-write `hits` every time.)
```{r}
hits = which(character_metadata$Gender == "M")
men = character_metadata$Character.ID[hits]
men[1:5]
```

So... what percentage of words in Shakespeare are spoken by male characters?

```{r}
hits = which(colnames(C) %in% men)
sum(C[,hits]) / sum(C)
```

This total, 76%, is actually an understatement. You'll see when we dig into the `character_metadata` information that many characters are left without a listed gender. In some cases that's an omission on our part. In others it's because we decided not to gender presumptively male minor figures, like messengers. 



## YOUR ASSIGNMENT ##

Choose any word (besides "death") in Shakespeare's corpus and report on the Homework Solutions forum six key facts:
1. What word are you looking at?
2. How many plays does it appear in?
3. How many speeches does it appear in?
4. How many characters say it at least twice?
5. How many times, total, is it said by female characters?
6. How many times, total, is it said in the comedies?

Honorable mentions will go to students who find the most interesting words that are used at least once by a female character. 

Super special honorable mentions will go to students who post a request for help with this assignment and/or respond with help.

